---
jupyter:
  jupytext:
    text_representation:
      extension: .Rmd
      format_name: rmarkdown
      format_version: '1.2'
      jupytext_version: 1.3.2
  kernelspec:
    display_name: R
    language: R
    name: ir
---

```{r}
#Install needed packages
#install.packages("docstring")
```

```{r}
library(docstring)
```

```{r}
# Sample dataset D
# (x1, x2, x3) are binary variables
x1 <- c(1, 1, 0, 1, 0, 0, 1, 0, 1, 0)
x2 <- c(0, 1, 0, 1, 0, 1, 1, 0, 1, 0)
x3 <- c(0, 1, 1, 1, 0, 1, 1, 0, 1, 0)
df <- data.frame(x1, x2, x3)
```

```{r}
log.fact <- function(m, r){
    #' Precompute log of factorials from 0 to m+r-1
    #'
    #' `log.fact` returns an array `a` of m+r elements, such that `a[i+1] = log(i!)`
    #'
    #' @param m Integer
    #' @param r Integer
    #' @return Numeric vector of size `m+r`, such that the (i+1)-th element contains log(i!).
    #'
    #' @examples
    #' log.fact(2, 3)
    
    fact <- log( c(1, 1:(m+r-1)) )

    for (i in 4:length(fact)) { #0! = 1, 1! = 1, 2! = 2, so they are already correct
        fact[i] <- fact[i] + fact[i-1]
    }
    return(fact)
}
```

The following function computes:
$$f(i, \mathbf{\pi_i}) = \prod_{j=1}^{q_i} \frac{(r_i - 1)!}{(N_{ij} + r_i - 1)!} \prod_{k=1}^{r_i} N_{ijk} $$
where:
- $r_i$ is the number of values that the $i$-th variable ($x_i$) can assume (e.g. $x_i \in \{0,1\} \Rightarrow r_i = 2$
- $\mathbf{\pi_i}$ contains the indices of the parents of $i$ (e.g. if $x_2, x_3 \rightarrow x_i$ in the DAG, then $\mathbf{\pi_i} = (1,2)$)
- $q_i$ is the number of possible values that the variables in $\mathbf{\pi_i}$ can take. For example, if $\mathbf{\pi_i} = (1,2)$, and both $x_1$ and $x_2$ are binary, then $q_i = 2^2$, since $(x_1, x_2) \in \{(0,0), (0,1), (1,0), (1,1)\}$. Treating the values of $x_j \in \mathbf{\pi_i}$ as digits of a binary number (or, in general, of a number in base $r = \max_j(r_j)$) defines a natural ordering for the possible values of $\mathbf{\pi_i}$.
- $N_{ijk}$ is the number of cases in the database $D$ where $x_i$ assumes its $k$-th value, and $\mathbf{\pi_i}$ its $j$-th value.
- $N_{ij} = \sum_{k=1}^{r_i} N_{ijk}$

```{r}
f <- function(i, i.parents, database, r, factorials) {
    #' Computes the function f
    #'
    #' Returns the joint probability that node i has parents i.parents, and that this bayesian network 
    #' has produced all the cases observed in database.
    #'
    #' @param i Integer. Represents a node.
    #' @param i.parents Vector of integers. Parents of i. Should not contain i itself.
    #' @param database data.frame of size (m,n) with all integer values between 0 and r-1. Observed cases.  
    #' @param r Integer. Number of possible values a single observation in database can assume.
    #' @param factorials Numeric vector. Precomputed log(factorial(i)) for i from 0 to m+r-1 (see log.fact function)
    #' @return Numeric. Probability that i has i.parents as parents, and that this structure results
    #' in the observed database.
    
    parents.length <- length(i.parents)
    m <- nrow(database)
    
    #From each row of database we compute the indices j and k
    
    #k: value of the i-th variable (database[i]) + 1
    #+1 is needed because indices in R start from 1, and database contains values 0:r-1
    index.k <- database[i] + 1
    
    #j: take values of database[i.parents], and treat them as digits of a number in base r 
    #(then add 1 to account for R indices)
    todec <- r ** (0:max(0,parents.length-1)) 
    
    if (parents.length > 0)
        index.j <- as.matrix(database[i.parents])%*%todec + 1
    else 
        index.j <- rep(1,m)

    #Gather the indices in a matrix
    indexes <- cbind(index.j, index.k) 
    
    #Sort
    indexes <- indexes[ order(indexes[,1], indexes[,2]),  ]
    
    #Now indexes is a sequence of "ordered blocks of rows", e.g.
    # indexes <- [[1, 1], [1, 2], [1, 2], [2, 1], [2, 2]]
    #We want to count the length of each block (i.e. number of "repetitions" of each row)
    #By taking the differences between each row and the previous one:
    d <- abs( diff( as.matrix(indexes) ) )
    #the only rows with non-zero elements are the ones at the "boundaries" of blocks
    
    #The example from before leads to:
    #d <- [[0, 1], [0, 0], [1, 1], [0, 1]]
    
    #Sum columns to reduce the matrix to a vector:
    d1 <- d[,1] + d[,2]
    #d1 <- [1, 0, 2, 1]
    
    #The unique rows are the ones with non-zero values + the last one
    idx.unique <- indexes[c(which(d1>0), length(d1)+1), ]
    
    #To compute repetitions, we add an element at the boundaries (start and end)
    #and take the differences of the indices of non-zero values
    N_ijk <- diff( c(0,which(d1>0), length(d1)+1) )
    
    #Gather everything in a data.frame
    N <- data.frame('Row'=idx.unique[,1], 'Col'=idx.unique[,2], 'Value'=N_ijk)
    #print(N)
    
    #Sum the elements of each row to compute N_{ij}
    Nj <- aggregate(.~Row,data=N,sum)[,c(3)]
    
    #Compute factorials
    N$Value <- factorials[N$Value+1]  
    logN <- aggregate(.~Row,data=N,sum)[,c(3)]
    
    #Compute the probability
    factor.first  <- factorials[r]    #r_i-1+1
    factor.second <- factorials[Nj+r] #+r_i -1 +1 (because of the indices)
    factor.third  <- logN

    result <- sum(factor.first - factor.second + factor.third)

    return(result)
}
```

```{r}
# K2 algorithm implementation

# Parameters:
# - database : 
# - u : integer
#   Maximum number of parents for each node.
# - ordering : 
# Output:
# adj : 

K2 <- function(database, u, ordering) {   
    #' Implementation of K2 algorithm as specified by 
    #' G. F. Cooper, E. Herskovits '"A Bayesian Method for the Induction of Probabilistic Networks from Data"
    #'
    #' Given a `database` of cases each containing n observations, and an `ordering` of the n variables,
    #' the algorithm finds the bayesian network structure with the (approximately) maximum probability
    #' of generating the observed data, by using at most `u` parents for each node i, which are chosen
    #' only between the nodes j preceding i in the ordering.
    #' 
    #' @param database data.frame of size (m, n). Dataset containing m observations of n variables, with no missing values.
    #' All values must be integers between 0 and r-1.
    #' @param u Integer. Maximum number of parents for each node.
    #' @param ordering Vector of size n OR list of n1 vectors of total size n
    #' If vector must contain a valid permutation of the integers [1, 2 ... n], representing the "prior" ordering of variables,
    #' such that the variable with index ordering[i] can have as possible parents only variables with index ordering[j] with j < i
    #' (i.e. variables that "appear before it" in the provided ordering).
    #' If list must contains the layers of the networks, representing the "prior" ordering of variables,
    #' such that the variable with index ordering[[i]] can have as possible parents only variables in the previous layers, and so
    #' with index ordering[[j]] with j<i. Notice that n1<=n, and if it is equal we come back to the previous case. 
    #' @return matrix of size (n, n)
    #' Adjacency matrix adj of the most probable Directed Acyclic Graph found given the evidence in the database.
    #' adj[i,j] is 1 if there is a connection from node j to node i (i.e. if j is a parent of i) 
    
    m <- nrow(database)
    n <- ncol(database)
    
    r <- max(database) + 1
    factorials <- log.fact(m, r)
    
    adj <- matrix(data = 0, nrow = n, ncol = n)
    
    debug <- TRUE
    
    for (i in 1:n) {
        i.parents <- c()
        
        log.p.old <- f(i, i.parents, database, r, factorials)
        #log-Probability of a structure with i disconnected
        
        if (debug) cat('x', i, ' [], log(p) = ', log.p.old, '\n')
        
        OKToProceed <- TRUE
        
        #Determine which nodes can be parents of i
        if (class(ordering)=='list'){ # Layer ordering, each element of the list is a layer
            
            for (h in 1:length(ordering)){
               if (i %in% ordering[[h]] ){
                   i.pos <- h
                   break
               } 
            }
            if( i.pos == 1){
                next
            } else {
                i.parents.candidates <- unlist( ordering[1:i.pos-1] )
            }
        } else { # Normal ordering, the input is a vector
            i.pos <- which(ordering == i) #Position of i in the ordering. Only variables before this position can be parents of i
        
            if (i.pos == 1) { #If i is at the start of the ordering, there are no parents to check
                next 
            } else {
                i.parents.candidates <- ordering[1:i.pos-1]
            }
        }
    
        #Add at most u nodes as parents of i,
        #only if each one of them increases the structure's probability
        while (OKToProceed & length(i.parents) < u) {
            log.p.new <- -Inf
            newparent.index <- NA
            
            #Select the node from the candidate list that maximizes the structure's probability (greedy algorithm)
            for (candidate in i.parents.candidates) {
                log.p.candidate <- f(i, c(i.parents, candidate), database, r, factorials)
                
                if (debug) cat('x', i, ' [', i.parents, ', +', candidate, '+], log(p) = ', log.p.candidate, '\n')
                
                if (log.p.candidate > log.p.new) {
                    log.p.new <- log.p.candidate
                    newparent.index <- candidate
                }
            }
        
            #Accept the new parent candidate only if it increases the total structure's probability
            if (log.p.new > log.p.old) {
                log.p.old <- log.p.new
                i.parents <- c(i.parents, newparent.index)
                
                if (debug) cat(newparent.index, ' -> ', i, '\n')
                adj[i, newparent.index] = 1
                
                #Remove the added parent from the candidates list
                i.parents.candidates <- i.parents.candidates[-newparent.index]
                
                if (length(i.parents.candidates) == 0) { #If there are no more candidates to check, terminate
                    OKToProceed <- FALSE
                    if (debug) cat('\n---\n')
                }
            } else { #If probability does not increase, stop adding parents to i
                OKToProceed <- FALSE
                if (debug) cat('\n---\n')
            }
        }
    }
    
    return(adj)
}
```

```{r}
#Print the structure
structure <- K2(df, 2, c(1,2,3))

print(structure)

#Should be:
# [0, 0, 0]
# [1, 0, 0]
# [0, 1, 0]

for (i in 1:nrow(structure)) {
    cat("Node ", i, " : [")
    
    for (j in 1:ncol(structure)) {
        if (structure[i,j]) {
            cat(j)
        }
    }
    
    cat("]\n")
}
```

**Expected output** (also checked by hand)
```text
x 1  [], log(p) =  -7.927324 
x 2  [], log(p) =  -7.927324 
x 2  [ , + 1 +], log(p) =  -6.802395 
1  ->  2 

---
x 3  [], log(p) =  -7.745003 
x 3  [ , + 1 +], log(p) =  -7.495542 
x 3  [ , + 2 +], log(p) =  -5.192957 
2  ->  3 
x 3  [ 2 , + 1 +], log(p) =  -5.991465 

---
     [,1] [,2] [,3]
[1,]    0    0    0
[2,]    1    0    0
[3,]    0    1    0
Node  1  : []
Node  2  : [1]
Node  3  : [2]```
